To be FAIR: Theory Specification Needs an Update

### 2025-02-19 manuscript.Rmd

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->

>>>>

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.

<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->
